Peptide retention time prediction

### 2015-01-07 retention.tex
x

Today a majority of the mass spectrometry (MS) protocols in use are
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms to interpret this data a major
area of proteomics research. A particular difficult challenge in this
context is posed by the tremendous complexity of the data. Proteins
display a wide range of physico-chemical properties and are often
found at very different amounts in complex samples, spanning as much
as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications~\cite{Lemeer2009}. To manage this
formidable diversity, increasing efforts are made to extract besides
the fragmentation spectra additional information from each MS
experiment.

An example of such additional information is represented by retention
times, i.e. the time points when peptides elute from the liquid
chromatography column.  The main purpose of liquid chromatography is
to reduce the number of peptides simultaneously injected into the mass
spectrometer. This increases the probability of each individual
peptide to be selected for fragmentation, which in its turn, increases
the proteome coverage achieved in an experiment.  However, besides
this, the separation process in itself provides useful information
about the analytes. On the one hand, the instrument always records the
time when a peptide elutes from the chromatography column, i.e. the
retention time. On the other hand, the retention time can also be
estimated by examining the sequence of the peptide. The availability
of both predicted and observed retention times has proven to be a
powerful feature, with several applications in MS-based proteomics
\cite{Palmblad2013}.

@@

\subsection{Machine learning-based methods}

In machine learning-based approaches a set of training peptides,
i.e. peptide sequences with known retention times, are used to
customize a predefined mathematical model. This step is typically
called training or learning. Once trained, the mathematical model can
be subsequently used to predict retention time for other peptides of
interest.

@@

Moruz {\em et al.} developed a software package for retention time
prediction which we named {\sc Elude} \cite{elude1, elude2}. One of
the main novelty points of {\sc Elude} was the calculation of peptide
features based on both a general hydrophobicity index, as well as a
retention index trained for each dataset at hand. This ensured that
the algorithm could be applied to different chromatography setups with
minimal loss in performance. For cases when little training data was
available, {\sc ELUDE} implemented an additional workflow based on a
library of pre-trained retention models. This workflow required only a
few tens of training peptides to select and calibrate the most
suitable model from the library. Both the selection and calibration
were performed using statistical methods robust to outliers.

@@

To address this, in our group we extended the first version of our
algorithm {\sc ELUDE} to encompass modified peptides as well. This was
straightforward to achieve once we observed that the in-house trained
retention index contained nearly all the information incorporated in
the more general hydrophobicity scale. This meant that we could remove
all the features based on this index with minimal loss in
accuracy. Further, we redefined all features describing a peptide in
terms of the in-house retention index alone. Since this index was
trained for each dataset at hand, the algorithm became fully trainable
for any modified amino acid (Figure \ref{fig:elude}). We evaluated the
new version of {\sc ELUDE} on peptides containing artifactual
modifications introduced as part of an MS protocol called COmbined
FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. Our
results showed that {\sc ELUDE} yielded equally accurate predictions
for modified and unmodified peptides \cite{elude2}.

@@

The current accuracy of the de novo retention time predictors that we
have discussed here is still much lower than the methods based on look
ups of previous registered retention time. The large availability of
data leads one to think that there still is a possibility to increase
the accuracy of such predictions by better technology. 
% Shorten the next two paragraphs
We can see that
there are a number of interesting machine learning techniques that not
yet have been used on the problem of predicting peptides retention
time. Here we will bring up one such techniques, that we feel are
particularly good matches, that we have not yet seen implemented for
the problem. Pairwise Machine-learned ranking (MLR)
\cite{liu2009learning} models can be trained to determine which of two
examples that should be better ranked thatn the other. MLR is
traditionally used for construction of ranking models for information
retrieval systems. However, in the case of retention time prediction
we can use such a predictor to predict the elusion order for sets of
peptides, which can be useful in its own \cite{bailey2012instant}, but
can also be converted into absolute retention times, given that the
distribution of their retention times are known.

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as bigdata, have put quite some
intest in so called deep learning techniques such as belief networks
using restricted multiple layered bolzman machines
(RBM) \cite{salakhutdinov2009deep} or deep neural networks. This
follows the observation that ideathe performance of predictors often
are enhanced if one can structure the prediction into multiple
layers. The drawback of such models are that they require large
training sets. Given the very large sets of peptides with retention
times that are assembled in various repositories, such as peptide
atlas and pride, such methods could probably be implemented for
retention time prediction as well.

Interpretation of mass spectrometry data is a question of
understanding the properties that explain the behavior of peptides in
a mass spectrometry experiment. The available technologies to infer
peptides and proteins from mass spectrometry data involve predicting
properties of peptides from a database, and compare those properties
to the observed data. Such comparisons are done for discovery
proteomics by algorithms called search engines, that traditionally are
limited to predicting predicting mass-to-charge rates of the precursor
peptides, and their fragmentation patterns. However, an attractive way
to increase the sensitivity of the inference procedure is to also
predict properties such as their iso-electric
point~\cite{branca2014hirief} or of particular interest for this
review, chromatographic retention time\cite{cerqueira2010mude}.

Moreover, targeted proteomics experiments, such as Selected Reaction
Monitoring (SRM) experiments, are dependent on the predictability of a
peptides behavior. In such experiments the retention times and the
suitable transitions are needed as input for the analysis. Frequently,
such information is derived from previous shotgun analysis. As the
capacity of targeted proteomics experiments expands by the
introduction of the more comprehensive data independent acquisition
techniques techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
pre-identification of peptide transitions and retention times
threatens to become a limiting factor. Our current methods to identify
peptides from DIA are only as comprehensive as the previous shotgun
analysis would be. An attractive alternative is to predict retention
time and transition using {\em de novo} predictions from the peptides
amino acid sequences.  Such a system of identifications not dependent
on previous peptide identifications from shotgun proteomics experiment
would lead to a more direct experimental design and hopefully to more
comprehensie coverage. Thus {\em de novo} retention time predictions
have an important role to fill.
%However, to do so we need to make the technique more accurate.

>>>>

Today a majority of the mass spectrometry (MS) protocols in use are
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms for data interpretation a major
area within proteomics research. A particular difficult challenge in
this context is posed by the tremendous complexity of the data.
Proteins display a wide range of physico-chemical properties and are
often found at very different amounts in complex samples, spanning as
much as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications~\cite{Lemeer2009}. To manage this
diversity, increasing efforts are made to extract besides the
fragmentation spectra additional information from each MS experiment.

An example of such additional information is represented by retention
times, {\em i.e.} the time points when peptides elute from the liquid
chromatography column.  The main purpose of liquid chromatography is
to reduce the number of peptides simultaneously injected into the mass
spectrometer. This increases the probability of each individual
peptide to be selected for fragmentation, which in its turn, increases
the proteome coverage achieved in an experiment.  However, besides
this, the separation process in itself provides useful information
about the analytes. On the one hand, the instrument always records the
time when a peptide elutes from the chromatography column, {\em i.e.}
the retention time. On the other hand, the retention time can also be
estimated by examining the sequence of the peptide. The availability
of both predicted and observed retention times has proven to be a
powerful feature, with several applications in MS-based proteomics
\cite{Palmblad2013}.

@@

\subsection{Machine learning-based methods}
\label{sec:ML}

In machine learning-based approaches a set of training peptides,
{\em i.e.} peptide sequences with known retention times, are used to
customize a predefined mathematical model. This step is typically
called training or learning. Once trained, the mathematical model can
be subsequently used to predict retention time for other peptides of
interest.

@@

Moruz {\em et al.} \cite{elude1} developed a software package for
retention time prediction which we named {\sc Elude}. One of the main
novelty points of {\sc Elude} was the calculation of peptide features
based on both a general hydrophobicity index, as well as a retention
index trained for each dataset at hand. This ensured that the
algorithm could be applied to different chromatography setups with
minimal loss in performance. For cases when little training data was
available, {\sc Elude} implemented an additional workflow based on a
library of pre-trained retention models. This workflow required only a
few tens of training peptides to select and calibrate the most
suitable model from the library. Both the selection and calibration
were performed using statistical methods robust to outliers.

@@

To address this, in Moruz {\em et al.}\cite{elude2} extended the first
version of their algorithm {\sc ELUDE} to encompass modified peptides
as well. This was straightforward to achieve once the authors observed
that the in-house trained retention index contained nearly all the
information incorporated in the more general hydrophobicity scale.
This meant that they could remove all the features based on this index
with minimal loss in accuracy. Further, they redefined all features
describing a peptide in terms of the in-house retention index alone.
Since this index was trained for each dataset at hand, the algorithm
became fully trainable for any modified amino acid (Figure
\ref{fig:elude}). The new version of {\sc ELUDE} was evaluated on
peptides containing artifactual modifications introduced as part of an
MS protocol called COmbined FRActional DIagonal Chromatography
(COFRADIC) \cite{Gevaert2002}. The results showed that {\sc ELUDE}
yielded equally accurate predictions for modified and unmodified
peptides \cite{elude2}.

@@

The accuracy of the machine learning-based methods described in
section \ref{sec:ML} is still much lower than the methods based on
look ups of previous registered retention time. However, the large
availability of datasets leads one to think that there still is a
possibility to increase the accuracy of such predictions by enhanced
machinelearning technology.  We can see that there are a number of
interesting techniques that not yet have been used on the problem of
predicting peptides retention time. Here we will bring up one such
techniques, that we feel are particularly good matches, that we have
not yet seen implemented for the problem. Pairwise Machine-learned
ranking (MLR) \cite{liu2009learning} models can be trained to
determine which of two examples that should be better ranked thatn the
other. MLR is traditionally used for construction of ranking models
for information retrieval systems. However, in the case of retention
time prediction we can use such a predictor to predict the elusion
order for sets of peptides, which can be useful in its own
\cite{bailey2012instant}, but can also be converted into absolute
retention times, given that the distribution of their retention times
are known.

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as bigdata, have put quite some
intest in so called deep learning techniques such as belief networks
using restricted multiple layered bolzman machines (RBM)
\cite{salakhutdinov2009deep} or deep neural networks. This follows the
observation that ideathe performance of predictors often are enhanced
if one can structure the prediction into multiple layers. The drawback
of such models are that they require large training sets. Given the
very large sets of peptides with retention times that are assembled in
various repositories, such as Peptide Atlas \cite{desiere2006peptideatlas} and
PRIDE \cite{martens2005pride}, such methods could probably be
implemented for retention time prediction as well.

The available technologies to infer peptides and proteins from mass
spectrometry data involve predicting the the potential manifestations
of the analyzed peptides, such as their fragmentation pattern, and
comparing those predicted properties to the observed data. Such
comparisons are regularly done for discovery proteomics by search
engines. However, an attractive way to increase the sensitivity of the
inference procedure is to also predict properties such as their
iso-electric point~\cite{branca2014hirief} or of particular interest
for this review, chromatographic retention
time\cite{cerqueira2010mude}.

Moreover, targeted proteomics experiments, such as Selected Reaction
Monitoring (SRM) experiments, are dependent on the predictability of a
peptides behavior. In such experiments the retention times and the
suitable transitions are needed as input for the analysis. Frequently,
such information is derived from previous shotgun analysis. As the
capacity of targeted proteomics experiments expands by the
introduction of the more comprehensive data independent acquisition
techniques techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
pre-identification of peptide transitions and retention times
threatens to become a limiting factor. Our current methods to identify
peptides from DIA are only as comprehensive as the previous shotgun
analysis would be. An attractive alternative is to predict retention
time and transition using {\em de novo} predictions from the peptides
amino acid sequences.  Such a system of identifications not dependent
on previous peptide identifications from shotgun proteomics experiment
would lead to a more direct experimental design and hopefully to more
comprehensie coverage. Thus {\em de novo} retention time predictions
have an important role to fill.
